---
date: 2017-03-29T06:45:48+02:00
title: "Make ggplot2 purrr"
tags: [R]
menu:
  main:
    parent: Blog
    identifier: /blog/ggplot2_purrr
    weight: 2
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggthemes)
library(dplyr)
library(tidyr)
library(purrr)
library(pwt9)
```

*Update*: I've included another way of saving a separate plot by group in this article, as pointed out
by [`@monitus`](https://twitter.com/monitus/status/849033025631297536). Actually, this is the preferred
solution; using `dplyr::do()` is deprecated, according to Hadley Wickham [himself](https://twitter.com/hadleywickham/status/719542847045636096).

I'll be honest: the title is a bit misleading. I will not use `purrr` that much in this blog post.
Actually, I will use one single `purrr` function, at the very end. I use `dplyr` much more.
However *Make ggplot2 purrr* sounds better than *Make ggplot dplyr* or whatever the verb for `dplyr` would be.

Also, this blog post was inspired by a stackoverflow question and in particular one of the
[answers](http://stackoverflow.com/a/29035145/1298051). So I don't bring anything new to the table,
but I found this stackoverflow answer so useful and so underrated (only 16 upvotes as I'm writing
this!) that I wanted to write something about it.

Basically the idea of this blog post is to show how to create graphs using `ggplot2`, but by
grouping by a factor variable beforehand. To illustrate this idea, let's use the data from the [Penn World
Tables 9.0](http://www.rug.nl/ggdc/productivity/pwt/). The easiest way to get this data is to
install the package called `pwt9` with:

```{r, eval=FALSE}
install.packages("pwt9")
```

and then load the data with:

```{r, eval=FALSE}
data("pwt9.0")
```

Now, let's load the needed packages. I am also using `ggthemes` which makes themeing your ggplots
very easy. I'll be making [Tufte](https://en.wikipedia.org/wiki/Edward_Tufte)-style plots.

```{r, eval=FALSE}
library(ggplot2)
library(ggthemes)
library(dplyr)
library(tidyr)
library(purrr)
library(pwt9)
```

First let's select a list of countries:

```{r}
country_list <- c("France", "Germany", "United States of America", "Luxembourg", "Switzerland", "Greece")

small_pwt <- pwt9.0 %>%
  filter(country %in% country_list)
```

Let's us also order the countries in the data frame as I have written them in `country_list`:

```{r}
small_pwt <- small_pwt %>%
  mutate(country = factor(country, levels = country_list, ordered = TRUE))
```

You might be wondering why this is important. At the end of the article, we are going to save the
plots to disk. If we do not re-order the countries inside the data frame as in `country_list`, the
name of the files will not correspond to the correct plots!

*Update*: While this can still be interesting to know, especially if you want to order the bars
of a barplot made with `ggplot2`, I included a suggestion by [`@expersso`](https://twitter.com/expersso/status/846986357792739328)
that does not require your data to be ordered!

Now when you want to plot the same variable by countries, say `avh` (*Average annual hours worked by
persons engaged*), the usual way to do this is with one of `facet_wrap()` or `facet_grid()`:

```{r}
ggplot(data = small_pwt) + theme_tufte() +
  geom_line(aes(y = avh, x = year)) +
  facet_wrap(~country)
```

```{r}
ggplot(data = small_pwt) + theme_tufte() +
  geom_line(aes(y = avh, x = year)) +
  facet_grid(country~.)
```

As you can see, for this particular example, `facet_grid()` is not very useful, but do notice its
argument, `country~.`, which is different from `facet_wrap()`'s argument. This way, I get the graphs
stacked horizontally. If I had used `facet_grid(~country)` the graphs would be side by side and completely
unreadable.

Now, let's go to the meat of this post: what if you would like to have one single graph for each
country? You'd probably think of using `dplyr::group_by()` to form the groups and then the graphs. This
is the way to go, but you also have to use `dplyr::do()`. This is because as far as I understand,
`ggplot2` is not `dplyr`-aware, and using an arbitrary function with groups is only possible with
`dplyr::do()`.

*Update*: As explained in the intro above, I also added the solution that uses `tidyr::nest()`:

```{r, eval = FALSE}
# Ancient, deprecated way of doing this
plots <- small_pwt %>%
  group_by(country) %>%
  do(plot = ggplot(data = .) + theme_tufte() +
       geom_line(aes(y = avh, x = year)) +
       ggtitle(unique(.$country)) +
       ylab("Year") +
       xlab("Average annual hours worked by persons engaged"))
```

And this is the approach that uses `tidyr::nest()`:

```{r}
# Preferred approach
plots <- small_pwt %>%
  group_by(country) %>%
  nest() %>%
  mutate(plot = map2(data, country, ~ggplot(data = .x) + theme_tufte() +
       geom_line(aes(y = avh, x = year)) +
       ggtitle(.y) +
       ylab("Year") +
       xlab("Average annual hours worked by persons engaged")))
```

If you know `dplyr` at least a little bit, the above lines should be easy for you to understand.
But notice how we get the title of the graphs, with `ggtitle(unique(.$country))`, which was actually
the point of the stackoverflow question.

*Update:* The modern version uses `tidyr::nest()`. Its documentation tells us:

*There are many possible ways one could choose to nest columns inside a data frame.
`nest()` creates a list of data frames containing all the nested variables: this seems to be the most useful form in practice.*
Let's take a closer look at what it does exactly:

```{r}
small_pwt %>%
  group_by(country) %>%
  nest() %>%
  head()
```

This is why I love lists in R; we get a `tibble` where each element of the column `data` is itself a `tibble`.
We can now apply any function that we know works on lists.

What might be surprising though, is the object that is created by this code. Let's take a look at
`plots`:

```{r}
print(plots)
```

As `dplyr::do()`'s documentation tells us, the return values get stored inside a list. And this is
exactly what we get back; a list of plots! Lists are a very flexible and useful class, and you cannot
spell *list* without `purrr` (at least not when you're a ne`R`d).

Here are the final lines that use `purrr::map2()` to save all these plots at once inside your working directory:

*Update*: I have changed the code below which does not require your data frame to be ordered according
to the variable `country_list`.

```{r, eval=FALSE}
# file_names <- paste0(country_list, ".pdf")

map2(paste0(plots$country, ".pdf"), plots$plot, ggsave)
```

As I said before, if you do not re-order the countries inside the data frame, the names of the files
and the plots will not match. Try running all the code without re-ordering, you'll see!

I hope you found this post useful. You can follow me on [twitter](https://www.twitter.com/brodriguesco) for
blog updates.

*Update*: Many thanks to the readers of this article and for their useful suggestions. I love the R
community; everyday I learn something new and useful!
