---
date: 2017-12-17
title: "Teaching the tidyverse to beginners"
tags: [R]
menu:
  main:
    parent: Blog
    identifier: /blog/teaching_tidyverse
    weight: 1
---

End October I tweeted this:

```{r, echo=FALSE}
blogdown::shortcode("tweet", "922741554992812032")
```

and it generated some discussion. Some people believe that this is the right approach, and some 
others think that one should first present *base* R and then show how the `tidyverse` complements
it. This year, I taught three classes; a 12-hour class to colleagues that work with me, a 15-hour
class to master's students and 3 hours again to some of my colleagues. Each time, I decided to 
focus on the `tidyverse`(almost) entirely, and must say that I am not disappointed with the results!

The 12 hour class was divided in two 6 hours days. It was a bit intense, especially the last 3 hours
that took place Friday afternoon. The crowd was composed of some economists that had experience 
with STATA, some others that were mostly using Excel and finally some colleagues from the IT
department that sometimes need to dig into some data themselves. Apart from 2 people, all the other
never had any experience with R. 

We went from 0 to being able to do the plot below after the end of the first day 
(so 6 hours in). Keep in mind that practically none of them even had opened RStudio before. I
show the code so you can see the progress made in just a few hours:

```{r, include=FALSE}
library(Ecdat)
library(tidyverse)
library(ggthemes)
```

```
library(Ecdat)
library(tidyverse)
library(ggthemes)
```

```{r}
data(Bwages)
bwages = Bwages %>%
  mutate(educ_level = case_when(educ == 1 ~ "Primary School",
                                educ == 2 ~ "High School",
                                educ == 3 ~ "Some university",
                                educ == 4 ~ "Master's degree",
                                educ == 5 ~ "Doctoral degree"))

ggplot(bwages) +
  geom_smooth(aes(exper, wage, colour = educ_level)) +
  theme_minimal() +
  theme(legend.position = "bottom", legend.title = element_blank())
```

Of course some of them needed some help here and there, and I also gave them hints (for example I told
them about `case_when()` and try to use it inside `mutate()` instead of nested ifs) 
but it was mostly due to lack of experience and because they hadn't had the time to fully digest 
R's syntax which was for most people involved completely new.

On the second day I showed `purrr::map()` and `purrr::reduce()` and overall it went quite well too. 
I even showed list-columns, and this is where I started losing some of them; I did not insist too 
much on it though, only wanted to show them the flexibility of `data.frame` objects. Some of them 
were quite impressed by list-columns! Then I started showing (for and while) loops and writing 
functions. I even showed them `tidyeval` and again, it went relatively well. Once they had the 
opportunity to play a bit around with it, I think it clicked (plus they have lots of code examples 
to go back too).

At the end, people seemed to have enjoyed the course, but told me that Friday was heavy; indeed it 
was, but I feel that it was mostly because 12 hours spread on 2 days is not the best format for this
type of material, but we all had time constraints.

The 15 hour Master's course was spread over 4 days, and covered basically the same. I just
used the last 3 hours to show the students some basic functions for model estimation 
(linear, count, logit/probit and survival models). Again, the students were quite impressed by how 
easily they could get descriptive statistics by first grouping by some variables. Through their 
questions, I even got to show them scoped versions of `dplyr` verbs, such as `select_if()` and
`summarise_at()`. I was expecting to lose them there, but actually most of them got these scoped 
versions quite fast. These students already had some experience with R though, but none with 
the `tidyverse`.

Finally the 3 hour course was perhaps the most interesting; I only had 100% total beginners. Some
just knew R by name and had never heard/seen/opened RStudio (with the exception of one person)!
I did not show them any loops, function definitions and no plots. I only showed them how RStudio 
looked and worked, what were (and how to install) packages (as well as the CRAN Task Views) and 
then how to import data with `rio` and do descriptive statistics only with `dplyr`. They were 
really interested and quite impressed by `rio` ("what do you mean I can use the same code for 
importing any dataset, in any format?") but also by the simplicity of `dplyr`.

In all the courses, I did show the `$` primitive to refer to columns inside a `data.frame`. First I 
showed them lists which is where I introduced `$`. Then it was easy to explain to them why it was 
the same for a column inside a `data.frame`; a `data.frame` is simply a list! This is also the 
distinction I made from the previous years; I simply mentioned (and showed really quickly) matrices
and focused almost entirely on lists. Most participants, if not all, had learned to program 
statistics by thinking about linear algebra and matrices. Nothing wrong with that, but I feel that R
really shines when you focus on lists and on how to work with them.

Overall as the teacher, I think that focusing on the `tidyverse` might be a very good strategy. I
might have to do some adjustments here and there for the future courses, but my hunch is that the
difficulties that some participants had were not necessarily due to the `tidyverse` but simply to 
lack of time to digest what was shown, as well as a total lack of experience with R. 
I do not think that these participants would have better understood a more traditional, `base`,
matrix-oriented course.
