---
date: 2018-04-10
title: "Get basic summary statistics for all the variables in a data frame"
tags: [R]
menu:
  main:
    parent: Blog
    identifier: /blog/brotools_describe
    weight: 1
---

I have added a new function to my `{brotools}` package, called `describe()`, 
which takes a data frame as an argument, and returns another data frame with descriptive
statistics. It is very much inspired by the [`{skmir}`](https://github.com/ropenscilabs/skimr)
package but also by 
[`assist::describe()`](https://github.com/bjornerstedt/assist/blob/master/R/describe.R) (click 
on the packages to be redirected to the respective Github repos)
but I wanted to write my own for two reasons: first, as an exercice, and second
I really only needed the function `skim_to_wide()` from `{skimr}`. So instead of installing a 
whole package for a single function, I decided to write my own (since I use `{brotools}` daily).

Below you can see it in action:

```{r, include=FALSE}
library(dplyr)
data(starwars)
```

```{r, eval=FALSE}
library(dplyr)
data(starwars)
```

```{r}
brotools::describe(starwars)
```

As you can see, the object that is returned by `describe()` is a `tibble`.

For now, this function does not handle dates, but it's in the pipeline. 

You can also only describe certain columns:

```{r}
brotools::describe(starwars, height, mass, name)
```

If you want to try it out,
you can install `{brotools}` from Github:

```
devtools::install_github("b-rodrigues/brotools")
```

If you found this blog post useful, you might want to follow me on [twitter](https://www.twitter.com/brodriguesco)
for blog post updates.
