---
date: 2018-12-27
title: "Some fun with {gganimate}"
tags: [R]
menu:
  main:
    parent: Blog
    identifier: /blog/fun_gganimate
    weight: 1
---

<div style="text-align:center;">
<video width="864" height="480" controls>
  <source src="/img/wiid_gganimate.webm" type="video/webm">
Your browser does not support the video tag.
</video>
</div>

In this short blog post I show you how you can use the `{gganimate}` package to create animations
from `{ggplot2}` graphs with data from UNU-WIDER.

## WIID data

Just before Christmas, UNU-WIDER released a new edition of their World Income Inequality Database:

```{r, echo=FALSE}
blogdown::shortcode("tweet", "1076001879556005888")
```

The data is available in Excel and STATA formats, and I thought it was a great opportunity to 
release it as an R package. You can install it with:

```{r, eval=FALSE}
devtools::install_github("b-rodrigues/wiid4")
```

Here a short description of the data, taken from UNU-WIDER's website:

*"The World Income Inequality Database (WIID) presents information on income inequality for 
developed, developing, and transition countries. It provides the most comprehensive set of income 
inequality statistics available and can be downloaded for free.*

*WIID4, released in December 2018, covers 189 countries (including historical entities), with over 
11,000 data points in total. With the current version, the latest observations now reach the year 2017."*

It was also a good opportunity to play around with the `{gganimate}` package. This package
makes it possible to create animations and is an extension to `{ggplot2}`. Read more about it 
[here](https://github.com/thomasp85/gganimate).

## Preparing the data

To create a smooth animation, I need to have a cylindrical panel data set; meaning that for each 
country in the data set, there are no missing years. I also chose to focus on certain variables
only; net income, all the population of the country (instead of just focusing on the economically
active for instance) as well as all the country itself (and not just the rural areas). 
On [this link](https://www.wider.unu.edu/sites/default/files/WIID/PDF/WIID4%20User%20Guide.pdf) you
can find a codebook (pdf warning), so you can understand the filters I defined below better.

Let's first load the packages, data and perform the necessary transformations:

```{r, eval=FALSE}
library(wiid4)
library(tidyverse)
library(ggrepel)
library(gganimate)
library(brotools)

small_wiid4 <- wiid4 %>%
    mutate(eu = as.character(eu)) %>%
    mutate(eu = case_when(eu == "1" ~ "EU member state",
                          eu == "0" ~ "Non-EU member state")) %>%
    filter(resource == 1, popcovr == 1, areacovr == 1, scale == 2) %>%
    group_by(country) %>%
    group_by(country, year) %>%
    filter(quality_score == max(quality_score)) %>%
    filter(source == min(source)) %>%
    filter(!is.na(bottom5)) %>%
    group_by(country) %>%
    mutate(flag = ifelse(all(seq(2004, 2016) %in% year), 1, 0)) %>%
    filter(flag == 1, year > 2003) %>%
    mutate(year = lubridate::ymd(paste0(year, "-01-01")))
```

For some country and some years, there are several sources of data with varying quality. I only
keep the highest quality sources with:

```{r, eval=FALSE}
    group_by(country, year) %>%
    filter(quality_score == max(quality_score)) %>%
```

If there are different sources of equal quality, I give priority to the sources that are the most
comparable across country (Luxembourg Income Study, LIS data) to less comparable sources with
(at least that's my understanding of the `source` variable):

```{r, eval=FALSE}
    filter(source == min(source)) %>%
```

I then remove missing data with:
 
```{r, eval=FALSE}
    filter(!is.na(bottom5)) %>%
```

`bottom5` and `top5` give the share of income that is controlled by the bottom 5% and top 5% 
respectively. These are the variables that I want to plot.

Finally I keep the years 2004 to 2016, without any interruption with the following line:

```{r, eval=FALSE}
    mutate(flag = ifelse(all(seq(2004, 2016) %in% year), 1, 0)) %>%
    filter(flag == 1, year > 2003) %>%
```

`ifelse(all(seq(2004, 2016) %in% year), 1, 0))` creates a flag that equals `1` only if the years
2004 to 2016 are present in the data without any interruption. Then I only keep the data from 2004
on and only where the flag variable equals 1. 

In the end, I ended up only with European countries. It would have been interesting to have countries
from other continents, but apparently only European countries provide data in an annual basis. 

## Creating the animation

To create the animation I first started by creating a static ggplot showing what I wanted; 
a scatter plot of the income by bottom and top 5%. The size of the bubbles should be proportional
to the GDP of the country (another variable provided in the data). Once the plot looked how I wanted
I added the lines that are specific to `{gganimate}`:

```{r, eval=FALSE}
    labs(title = 'Year: {frame_time}', x = 'Top 5', y = 'Bottom 5') +
    transition_time(year) +
    ease_aes('linear')
```

I took this from `{gganimate}`'s README.

```{r, eval=FALSE}
animation <- ggplot(small_wiid4) +
    geom_point(aes(y = bottom5, x = top5, colour = eu, size = log(gdp_ppp_pc_usd2011))) +
    xlim(c(10, 20)) +
    geom_label_repel(aes(y = bottom5, x = top5, label = country), hjust = 1, nudge_x = 20) +
    theme(legend.position = "bottom") +
    theme_blog() +
    scale_color_blog() +
    labs(title = 'Year: {frame_time}', x = 'Top 5', y = 'Bottom 5') +
    transition_time(year) +
    ease_aes('linear')
```

I use `geom_label_repel` to place the countries' labels on the right of the plot. If I don't do 
this, the labels of the countries would be floating around and the animation would be unreadable. 

I then spent some time trying to render a nice webm instead of a gif. It took some trial and error
and I am still not entirely satisfied with the result, but here is the code to render the animation:

```{r, eval=FALSE}
animate(animation, renderer = ffmpeg_renderer(options = list(s = "864x480", 
                                                             vcodec = "libvpx-vp9",
                                                             crf = "15",
                                                             b = "1600k", 
                                                             vf = "setpts=5*PTS")))
```

The option `vf = "setpts=5*PTS"` is important because it slows the video down, so we can actually
see something. `crf = "15"` is the quality of the video (lower is better), `b = "1600k"` is the 
bitrate, and `vcodec = "libvpx-vp9"` is the codec I use. The video you saw at the top of this 
post is the result. You can also find the video [here](link).

I would have preferred if the video was smoother, which should be possible by creating more frames.
I did not find such an option in `{gganimate}`, and perhaps there is none, at least for now.

In any case `{gganimate}` is pretty nice to play with, and I'll definitely use it more!

Hope you enjoyed! If you found this blog post useful, you might want to follow 
me on [twitter](https://www.twitter.com/brodriguesco) for blog post updates and 
[buy me an espresso](https://www.buymeacoffee.com/brodriguesco) or [paypal.me](paypal.me/brodriguesco).

<style>.bmc-button img{width: 27px !important;margin-bottom: 1px !important;box-shadow: none !important;border: none !important;vertical-align: middle !important;}.bmc-button{line-height: 36px !important;height:37px !important;text-decoration: none !important;display:inline-flex !important;color:#ffffff !important;background-color:#272b30 !important;border-radius: 3px !important;border: 1px solid transparent !important;padding: 1px 9px !important;font-size: 22px !important;letter-spacing:0.6px !important;box-shadow: 0px 1px 2px rgba(190, 190, 190, 0.5) !important;-webkit-box-shadow: 0px 1px 2px 2px rgba(190, 190, 190, 0.5) !important;margin: 0 auto !important;font-family:'Cookie', cursive !important;-webkit-box-sizing: border-box !important;box-sizing: border-box !important;-o-transition: 0.3s all linear !important;-webkit-transition: 0.3s all linear !important;-moz-transition: 0.3s all linear !important;-ms-transition: 0.3s all linear !important;transition: 0.3s all linear !important;}.bmc-button:hover, .bmc-button:active, .bmc-button:focus {-webkit-box-shadow: 0px 1px 2px 2px rgba(190, 190, 190, 0.5) !important;text-decoration: none !important;box-shadow: 0px 1px 2px 2px rgba(190, 190, 190, 0.5) !important;opacity: 0.85 !important;color:#82518c !important;}</style><link href="https://fonts.googleapis.com/css?family=Cookie" rel="stylesheet"><a class="bmc-button" target="_blank" href="https://www.buymeacoffee.com/brodriguesco"><img src="https://www.buymeacoffee.com/assets/img/BMC-btn-logo.svg" alt="Buy me an Espresso"><span style="margin-left:5px">Buy me an Espresso</span></a>

